Hierarchical Principal Manifold Estimation

### 2023-10-03 paper/main.tex
edits, added preliminary mnist results

The following sections are organized as follows. Section \ref{s:methods} specifies an additive smoothing spline model for use in the analysis of neuroimaging data from the ADNI study, and provides a comprehensive description of the hierarchical PME algorithm developed in this paper. Section \ref{s:hpme_mnist_simulation} applies this method to the MNIST dataset of handwritten digits, and compares the results to those produced using PME. The performance of the two approaches is then compared when they are applied to a series of simulated datasets of varying complexities. Section \ref{s:hpme_adni} discusses proposed work, which will demonstrate how the model specified in Section \ref{s:methods} may be used for the estimation of a manifold underlying the surfaces of subcortical structures of participants of the ADNI dataset, while Section \ref{s:hpme_discussion} provides a concluding discussion of the proposed algorithm and opportunities for further study.

@@

\section{Proposed Work} \label{s:hpme_adni}
Further work will be dedicated to testing the proposed approach in several settings. First, the algorithm will be tested with a simplified additive model on the MNIST dataset of handwritten images. This will provide an opportunity to test the method on a relatively small imaging dataset, serving as a proof of concept and indicating where problems may arise when the algorithm is used with more complex data structures. The method will next be applied to a series of simulated datasets that were generated to reflect the hierarchical structure present in the ADNI dataset. In both of these cases, the estimates of the hierarchical approach will be compared to alternative methods, with the fit of each method to the data being compared. Finally, the hierarchical approach will be applied to structural MRI data from the ADNI dataset. Specifically, the algorithm can be used to estimate the manifolds underlying the hippocampus of study participants. Of interest in this case will be a comparison of the varying-coefficient models by group, which may provide a visual confirmation that the algorithm's estimates are detecting the most relevant information included in the data.

\section{Discussion} \label{s:hpme_discussion}

>>>>

The following sections are organized as follows. Section \ref{s:methods} specifies an additive smoothing spline model for use in the analysis of neuroimaging data from the ADNI study, and provides a comprehensive description of the hierarchical PME algorithm developed in this paper. Section \ref{s:hpme_mnist_simulation} applies this method to the MNIST dataset of handwritten digits, and compares the results to those produced using PME. The performance of the two approaches is then compared when they are applied to a series of simulated datasets of varying complexities. Section \ref{s:hpme_adni} discusses proposed work, which will demonstrate how the model specified in Section \ref{s:methods} may be used for the estimation of a manifold underlying the surfaces of subcortical structures of participants of the ADNI dataset. This will be followed by a concluding discussion of the proposed algorithm and opportunities for further study.

@@

To demonstrate the potential benefits of the proposed hierarchical estimation approach when compared to the existing PME algorithm, we implemented the hierarchical PME algorithm using a simplified version of the additive spline model presented in Section \ref{s:methods} for use on images from the MNIST database. The MNIST database includes 70,000 $28 \times 28$ images of handwritten numerical digits written by different individuals, with 60,000 in a training set and an additional 10,000 in a test set. To test the hierarchical algorithm, we considered a reduced set of 100 images of the numeral ``5", where the individual pixels of each image with intensity greater than 0 can be considered to create a 2-dimensional data cloud, through which a 1-dimensional curve passes. Each representation of the numeral must follow the same general path to be recognizable as a ``5", but differences in individual handwriting styles result in variation between images. We have no meaningful way of grouping images by individual or subgroup in this case, and there are no temporal relationships between images, so we assume the embedding function used in Algorithm \ref{alg:hpme} takes the form \[%
  \mathbf{x}_{jq} = \psi(\mathbf{y}_{jq}) = f_p(\mathbf{y}_{jq}) + f_j(\mathbf{y}_{jq}) + \epsilon_{jq}
.\]%

Plots depicting how the population-level manifold estimate differs from the manifolds estimated for each image are shown for four images in Figure \ref{fig:mnist_results}. Here, we see that the population estimate, represented in red, clearly captures the shape of a numeral ``5" and does not show any clear bias compared to the data points included in the plots. Meanwhile, the estimates for the individual images, shown in blue, are more responsive to the image data to which it is fit, indicating that the additive model fit successfully to the data, and providing a simple demonstration of the improvements made available by incorporating hierarchical information into the manifold learning procedure.

\begin{figure}%
  \centering
  \subfloat[\centering MNIST Image 1]{{\includegraphics[width=9cm]{mnist_5plot1}}}%
  \hfill
  \subfloat[\centering MNIST Image 2]{{\includegraphics[width=9cm]{mnist_5plot2}}}
  \vfill
  \subfloat[\centering MNIST Image 3]{{\includegraphics[width=9cm]{mnist_5plot3}}}
  \hfill
  \subfloat[\centering MNIST Image 4]{{\includegraphics[width=9cm]{mnist_5plot4}}}
  \caption{Comparison of PME, Population, and Image-Level Results for MNIST Data}
  \label{fig:mnist_results}
\end{figure}

\section{Proposed Work} \label{s:hpme_adni}
Further work will be dedicated to testing the proposed approach in several settings. First, more extensive performance comparisons between the PME algorithm and its hierarchical extension will be conducted when applied to the MNIST dataset. The method will next be applied to a series of simulated datasets that were generated to reflect the hierarchical structure present in the ADNI dataset. Again, the estimates of the hierarchical approach will be compared to alternative methods, with the fit of each method to the data being compared. Finally, the hierarchical approach will be applied to structural MRI data from the ADNI dataset. Specifically, the algorithm can be used to estimate the manifolds underlying the hippocampus of study participants. Of interest in this case will be a comparison of the varying-coefficient models by group, which may provide a visual confirmation that the algorithm's estimates are detecting the most relevant information included in the data.
